---
title: "Precipitation"
author: "Erika W"
date: "3/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Precipitation

### Data

In October 2020, Steven Eikenbary created a 4x4km grid of the SF delta study area in ArcMap, and retrieved the lat/longs of the center points in each grid. Those coordinates can be found at:

URL

On October 26, 2020, precipitation data was downloaded from PRISM for that grid of coordinates (n=33) using the multiple sites downloader: https://prism.oregonstate.edu/explorer/bulk.php

Data were downloaded for every water year between 2009 through 2019, for all locations within the grid.

The spatial resolution of the data is 4km, and it was sourced from the AN81d dataset. Precipitation includes rainfall + melted snowfall. Additional details about the data can be found at: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf

```{r}
df2009 <- read.csv("")

```

### Filtering

Once I had that I filtered by daily max and then ran a window looking at when precipitation for a region was greater than 0.25 inches and >0.5 inches for a 24 hour period and then put that signifier on the data points that followed 'wet' precipitation events by 7 days.

The PRISM data manipulation starts around line 195 in the attached R file. It was pretty simple, but I'm sure I'm not doing a great job of explaining it here.
